---
title: "Group-6"
author: "Mathanki Mehra"
date: "2024-10-08"
output: html_document
---
# PART 1

To download a file from GitHub to R, the download.file() function can be employed. The following steps outline the procedure:

1. **Find the Raw File URL**:
   - Navigate to the desired file within the GitHub repository.
   - Click on the file and then select the "Raw" button to obtain the direct link to the file.

2. **Use download.file() in R**:
   - Utilize the download.file() function, specifying the URL of the raw file as the first argument and the destination path on the local machine as the second argument.

```{r}
# URL of the first raw file on GitHub
url <- "https://raw.githubusercontent.com/ghazkha/Assessment4/refs/heads/main/gene_expression.tsv"

# Path to save the file locally (in the current directory)
destination <- "gene_expression.tsv"

# Download the file
download.file(url, destination)

# URL of the second raw file on GitHub
url1 <- "https://raw.githubusercontent.com/ghazkha/Assessment4/refs/heads/main/growth_data.csv"

# Path to save the file locally (in the current directory)
destination <- "growth_data.csv"

# Download the file
download.file(url1, destination)

```

## Step 1
To read the gene_expression.tsv file into R and set the gene identifiers as the row names, the read.table() function can be used. Subsequently, the first six genes can be displayed using the head() function. The following code illustrates these steps:
```{r}
# Read in the gene expression data and making gene identifiers as row names
gene_expression <- read.table("gene_expression.tsv", header = TRUE, row.names = 1, sep = "\t")

# Show a table of values for the first six genes
head(gene_expression)
```

**Code Explanation:**

- header = TRUE: This tells R that the first row of the file contains the column names (i.e., headers for each column). This allows R to correctly assign the names to the data frame's columns.

- row.names = 1: This argument specifies that the first column of the data file (index 1) should be treated as row names instead of a regular data column. This means that the values in the first column will not be included as part of the data frame; instead, they will serve as labels for the rows.

- sep = "\t": This indicates that the data is tab-separated. Since the file is a .tsv (tab-separated values), this ensures that R reads the file correctly by treating each tab as a separator between columns.

By using the row.names = 1 argument, the gene identifiers from the first column are set as the row names of the gene_expression data frame. As a result, when we display the data frame using head(gene_expression), we will see the gene identifiers as the row labels.

We can also see the structure of the data frame after reading it, using str() command.

```{r}
str(gene_expression)
```

This shows the column names and the row names, allowing us to confirm that the gene identifiers are indeed set as row names. 

## Step 2
The following code is used to calculate the mean of columns as well as add a new column "mean_count" to the existing data frame. 

```{r}
# Read in the gene expression data
gene_expression <- read.table("gene_expression.tsv", header = TRUE, row.names = 1, sep = "\t")

# Calculate the mean of the counts across the columns for each gene
gene_expression$mean_count <- rowMeans(gene_expression)

# Show a table of values for the first six genes, including the new mean column
head(gene_expression)

```

**Code Explanation:**

- rowMeans(gene_expression): This function calculates the mean of the counts for each row (gene) across all the specified columns. The result is a vector containing the mean counts for each gene.

- gene_expression$mean_count <-: This assigns the calculated mean counts to a new column called mean_count in the gene_expression data frame.

- head(gene_expression): This displays the first six genes along with all their corresponding count data and the new mean column.

We can further use the View() command to visualize the data frame. By using this command we can further see the newly added column.

## Step 3
To identify the 10 genes with the highest mean expression from the gene_expression data frame, the order() function can be utilized to sort the data based on the mean_count column in descending order. Subsequently, the top 10 rows can be extracted. The following steps outline this process:
```{r}
# Read in the gene expression data
gene_expression <- read.table("gene_expression.tsv", header = TRUE, row.names = 1, sep = "\t")

# Calculate the mean of the counts across the columns for each gene
gene_expression$mean_count <- rowMeans(gene_expression)

# Get the 10 genes with the highest mean expression
top_10_genes <- gene_expression[order(-gene_expression$mean_count), ][1:10, ]

# Display the top 10 genes
top_10_genes

```
**Code Explanation:**

- order(-gene_expression$mean_count): This orders the indices of gene_expression based on the mean_count column in descending order (the negative sign - indicates descending order).

- gene_expression[...]: This uses the ordered indices to subset the gene_expression data frame, sorting the rows based on mean expression.

- [1:10, ]: This selects the first 10 rows of the sorted data frame, which correspond to the genes with the highest mean expression.

- top_10_genes: This stores the resulting data frame containing the top 10 genes, and the final line displays it.

## Step 4
The next step is to estimate the number of genes with a mean that is less than 10.This can be done with the help of the running the codes given below 
```{r}
#calculate the mean counts of the genes given in the list 
gene_expression$mean_count <- rowMeans(gene_expression)
```
**code explanation:**
The code rowMeans(gene_expression)is used to find the mean of all the values in a row for each gene given and this is done across all the columns or the genes present in the data.

gene_expression$mean_count is the code which directs the production of a new category in the data frame by the name of mean_count and all the dta produced by the code rowMeans(gene_expression) is shown under this particular category
```{r}
#find the mean count find that have mean values less than 10 
gene_expression$mean_count < 10
```
**code description:**
This code functions to find out the genes from the given data frame (gene_expression)which have mean values which are less than 10.with this code we can specifying to find the mean values that are less than 10 in the mean_count column  that is created in the gene_expression data frame ,and then represent it in the form of true/ false.
"true" represents that mean value is less than 10
"False " value means that the mean value is more than 10.

```{r}
# Count the number of genes_expression with mean_count less than 10
 sum(gene_expression$mean_count < 10, na.rm = TRUE)
```
**code explanation:**
To determine the number of genes that have mean values less than 10, we so the sum total of the categories that is done with the help of the code given above, which is 35988.


#Step 5
This step involves the construction of the histogram which is a bar graph representation of the mean expression values .

```{r}
#Retreive all the mean count values given in the mean_count column
gene_expression[,c("mean_count")]
```
**code explanation:**
the code given above is used to specify the rows and columns that are being considered.according to the code , we are considering the all the rows and the specific column of mean_count which represents the mean expression values.


```{r}
# histogram of mean gene expression values
hist(gene_expression$mean_count,xlab="mean expression value", main="Histogram of Mean Gene Expression Values")
```
**code explanation:**
The code given above is used to generate a histogram of the mean expression values where in the code the function "hist" is used to produce a histogram for the specified data , that is, gene_expression. in the code the main heading for the graph and the x- axis headings are also specified.

#Step 6
```{r}
#Import the csv file
growth_data <- read.csv("growth_data.csv")
```
**code explanation**
The code "read.csv" given above is used to import the csvfile named growth_data.csv and converts it into R object , where it is saved as growth_data.
"<-" function is used to direct the output of the read.csv function to the growth_data.

```{r}
#Display the column names
colnames(growth_data)
```
**Code explanation:**
the code given above is used to list the column names and that is done with the help of the function "colnames" after which we also specify the name of the data frame ,that is growth_data.


# selects all data from all rows in columns 3 to 6 and calculates the mean
colMeans(growth_data[, 3:6])

# selects all data from all rows in columns 3 to 6 and calculates the standard deviation, the 2 refers to column selection and apply() is used to apply the sd function to each column

apply(growth_data[, 3:6], 2,sd)

#code used to create a boxplot from columns 3 to 6
boxplot(growth_data[, 3:6])

#code combines columns 4, 5 and 6 and calculates the mean
mean(c(growth_data[,4], growth_data[,5], growth_data[,6]))



